A Database Management System for a Geospatially-Enabled Interactive Game

### 2018-10-28 paper/MSDS7330DBMSforGeospatiallyEnabledGame.tex
Changes in the paper

\section{Introduction}
Location based apps and services are becoming more popular and more popular. The game Pokémon Go, in particular, has gained over 44 million users and additionally incorporates augmented reality. We plan to develop a database that can collect, store and aid in processing information for a game where players will be collecting items in different tourist destinations to earn points. The game will consist of players moving around the real world and collecting tokens that gradually reveal parts of a story. The data will be collected from a user interface operating on their cellphones. The emphasis is to design a system that can efficiently collect and store geospatially enabled data, and use a database structure to manage the game rules and possible interactions among players.

\section{Game rules and design considerations}
\subsection{Problem statement}
As mentioned, the game will consist of players exploring the real world in search of tokens. Collecting tokens raises the player’s score which will in turn reveal a segment of a story once the player’s score reaches a certain threshold. The tokens belong to different categories that correspond to different scores that must be earned and different aspects of the story that may be revealed. As such, this game relies heavily on the search-and-find mechanic as well as desires for collection, completion, and closure of narrative for player engagement \cite{game-methodology, location-based-games}.

@@

\begin{enumerate}
	\item Nature of GPS Data. Learning GPS data types, how they are stored, and how they may be obtained both from the player and by the player.
	\item Available DBMS. What DMBS and existing software packages work well with GPS data and are also capable of implementing game rules.
	\item Computations GPS Data. What computations can be done with GPS data to enforce game rules and accomplish game mechanics such as the player interacting with geospatial objects.
	\item Database Schema. What entities and attributes are necessary to store and what portions of the game may be saved by the player.
\end{enumerate}

\subsection{Previous work related to the problem}
An interesting paper by Chulmo Koo and others investigate relationship between geospatially enabled mobile game and destination satisfaction \cite{destination-engagement}. Besides, some authors argue that location-based games are positively associated with a set of beneficial health behaviors \cite{pokemon-motivation}. We found it interesting that a game can have a positive effect on tourism and person’s wellness and would like to extend research in this area.  
   Initial research in this area brought us to a paper summarizing design patterns possible with location-based games \cite{location-based-mg}. Due to time and resource limits imposed on us we decided to use on Search-and-Find pattern for our game. On the other hand, the large-scale nature of multiplayer games might generate enormous data sets. Effective handling of this data requires a sophisticated database management system \cite{data-store-issues, location-based-services}. In his thesis, Kristian Midtgård looks at the NoSQL landscape in an attempt to find the best practices and candidate databases for achieving high write throughput \cite{massive-amounts-location-data}. We plan to build on top of his research and look at other, more modern databases.

@@

\begin{enumerate}[label=\Alph*]
\item REST over HTTP. 
\item SQL 
\end{enumerate}

@@

An important aspect when dealing with GPS location data is how to represent the data efficiently in order to reduce network traffic and storage requirements. A geospatial location is defined by its latitude and longitude coordinates. There are different formats in which these coordinates can be represented. The most common format for many applications (such as Google Maps and Geographical Information Systems or GIS) is the signed degrees format. The rules for this formatting are as follows:
\begin{itemize}
\item Latitudes range from -90 to 90, and longitudes from -180 to 180. 
\item South latitudes and West longitudes are preceded by a minus sign.
\item The format uses decimal places, which give the desired amount of precision. 
\end{itemize}

@@

This information is important because it will determine, along with the number of players, how much data is stored in the database. If the amount is very large, some sort of data compression would be necessary.

@@

\item MySQL: As explained above, a MySQL database will be able to support geospatial data, using either a numeric or a spatial format. The major advantages of using it would be our own familiarity with the interface, and its open source nature. A major limitation could be that it only supports B+ Trees, which could impose restrictions to the storage and indexing of the data, an important component of the project.

\item PostgreSQL: PostGIS is a spatial database extender for PostgreSQL object-relational database. It adds support for geographic objects allowing location queries to be run in SQL. The three features that PostGIS delivers to PostgreSQL DBMS are spatial types, indexes and functions such as distance, area, union, intersection, and specialty geometry. This spatial database stores and manipulates spatial objects like any other object in the database. With support for different geometry types, the PostGIS spatial database allows querying and managing information about locations and mapping.

@@

We will build a database based on the open-sourced PostgreSQL database, integrating the capabilities of the PostGIS extension. Apart from the aforementioned advantages of using these tools, our research indicates that PostGIS is close to becoming the standard to manage geolocation data for SQL databases.  This will allow us to build a database with the following capabilities:
\begin{itemize}
\item	Once the data is uploaded, use simple SQL language to run complex queries related locations, objects, and other spatial data. 
\item Scale out the database in various ways: a. Manage a very large data size because it offers multi-TB capabilities, b. Handle a large amount of concurrency from the data, c. Use table-partitioning to break large data sets down into more manageable pieces.
\item Build custom functions using a language such as Python or R, which will be very useful to program the game requirements, rules and mechanisms into the database. 
\end{itemize}

@@

Player information, story segment information, and information about the five categories of tokens can be stored in their own tables. There will be a central table that links these various entities via a Game-ID. This in anticipation that we may want to tell two stories in the same city. A list of geospatial points that correspond to viable latitudes and longitudes that can be used for any particular game will be stored separately from the list of geospatial points that are actually available to the player. This will allow us to update the list and properties of physical geospatial points without affecting existing games and applications. A diagram of the database schema can be found below in \autoref{ERD}.

>>>>

\section{Introduction}
Location based apps and services are becoming more and more popular. The game Pokémon Go, in particular, has gained over 44 million users and additionally incorporates augmented reality. We developed a database that can collect, store and aid in processing information for a game where players are collecting items in different tourist destinations to earn points. The game consist of players moving around the real world and collecting tokens that gradually reveal parts of a story. The data is collected from a user interface operating on their cellphones. The emphasis of this work is to design a system that can efficiently collect and store geospatially enabled data, and use a database structure to manage the game rules and possible interactions among players.

\subsection{Previous work related to the problem}
An interesting paper by Chulmo Koo and others investigate relationship between geospatially enabled mobile game and destination satisfaction \cite{destination-engagement}. Besides, some authors argue that location-based games are positively associated with a set of beneficial health behaviors \cite{pokemon-motivation}. We found it interesting that a game can have a positive effect on tourism and person’s wellness and would like to extend research in this area.  
   Initial research in this area brought us to a paper summarizing design patterns possible with location-based games \cite{location-based-mg}. Due to time and resource limits imposed on us we decided to use on Search-and-Find pattern for our game. On the other hand, the large-scale nature of multiplayer games might generate enormous data sets. Effective handling of this data requires a sophisticated database management system \cite{data-store-issues, location-based-services}. In his thesis, Kristian Midtgård looks at the NoSQL landscape in an attempt to find the best practices and candidate databases for achieving high write throughput \cite{massive-amounts-location-data}. We've built our work on top of his research and look at other, more modern databases.

\section{Game rules and design considerations}
\subsection{Problem statement}
As mentioned, the game consists of players exploring the real world in search of tokens. Collecting tokens raises the player’s score which will in turn reveal a segment of a story once the player’s score reaches a certain threshold. The tokens belong to different categories that correspond to different scores that must be earned and different aspects of the story that may be revealed. As such, this game relies heavily on the search-and-find mechanic as well as desires for collection, completion, and closure of narrative for player engagement \cite{game-methodology, location-based-games}.

@@

\begin{enumerate}
	\item Nature of Geospatial Data. Learning data types, how they are stored, and how they may be obtained both from the player and by the player.
	\item Available DBMS. What databases and existing software packages work well with GPS data and are also capable of implementing game rules.
	\item Using Location Data. What computations can be done with GPS coordinates to enforce game rules and accomplish game mechanics such as the player interacting with geospatial objects.
	\item Database Schema. What entities and attributes are necessary to store and what portions of the game may be saved by the player.
\end{enumerate}

@@

\begin{enumerate}[label=\Alph*]
\item REST over HTTP. This protocol is used to send messages from the client to the server sides and back
\item SQL. We use relational database and SQL over Application Layer to transform and query our data. 
\end{enumerate}

@@

An important aspect when dealing with geospatial data is how to represent the data efficiently in order to reduce network traffic and storage requirements. A geospatial location is defined by its latitude and longitude coordinates. There are different formats in which these coordinates can be represented. The most common format for many applications (such as Google Maps and Geographical Information Systems or GIS) is the signed degrees format. The rules for this formatting are as follows:
\begin{itemize}
\item Latitudes range from -90 to 90, and longitudes from -180 to 180. 
\item South latitudes and West longitudes are preceded by a minus sign.
\item The format uses decimal places, which give the desired amount of precision. 
\end{itemize}

@@

This information is important because it determines, along with the number of players, how much data is stored in the database. If the amount is very large, some sort of data compression would be necessary.

@@

\item MySQL: a MySQL database does not support geospatial data, using either a numeric or a spatial format. The major advantages of using this DBMS would be simplicity of the interface, and its open source nature. A major limitation could be that it only supports B+ Trees, which could impose restrictions to the storage and indexing of the data, an important component of the project.

\item PostgreSQL: PostGIS is a spatial database extention for PostgreSQL object-relational database. It adds support for geographic objects allowing location queries to be run in SQL. The three features that PostGIS delivers to PostgreSQL DBMS are spatial types, indexes and functions such as distance, area, union, intersection, and specialty geometry. This spatial database stores and manipulates spatial objects like any other object in the database. With support for different geometry types, the PostGIS spatial database allows querying and managing information about locations and mapping.

@@

Our work is based on the open-sourced PostgreSQL database, integrating the capabilities of the PostGIS extension. Apart from the aforementioned advantages of using these tools, our research indicates that PostGIS is close to becoming the standard to manage geolocation data for SQL databases.  This allows us to build a database with the following capabilities:
\begin{itemize}
\item	Once the data is uploaded, use simple SQL language to run complex queries related locations, objects, and other spatial data. 
\item Scale out the database in various ways: a. Manage a very large data size because it offers multi-TB capabilities, b. Handle a large amount of concurrency from the data, c. Use table-partitioning to break large data sets down into more manageable pieces.
\item Build custom functions using a language such as Python or R, which will be very useful to program the game requirements, rules and mechanisms into the database. 
\end{itemize}

@@

Player information, story segment information, and information about the five categories of tokens can be stored in their own tables. There will be a central table that links these various entities via a Game-ID. This in anticipation that we may want to tell two stories in the same city. A list of geospatial points that correspond to viable latitudes and longitudes that can be used for any particular game will be stored separately from the list of geospatial points that are actually available to the player. This allows us to update the list and properties of physical geospatial points without affecting existing games and applications. A diagram of the database schema can be found below in \autoref{ERD}.
